---
title: "Chapter 5 Marker Identification"
execute:
  warning: false
  message: false
format:
  html:
    embed-resources: true
---

```{r}
#| echo: false

# Load RData
load("data/chapter4.RData")
```

```{r}
# libraries for this chapter
library(tidyverse)
library(Seurat)
library(data.table)
library(HGNChelper)
```


# 1. Find marker genes

## 1.1 `FindAllMarkers`

This type of analysis is typically for evaluating a single sample group/condition. With the `FindAllMarkers` function, we are comparing each cluster to all other clusters to identify potential marker genes. The cells in each cluster are treated as replicates, and essentially a differential expression analysis is performed with some statistical test. By default, the test is Wilcoxon Rank Sum test.

The `FindAllMarkers` function has three arguments to determine if a gene is a marker gene.

-   `logfc.threshold` Limit testing to genes that show, on average, at least an X-fold difference (log-scale) between the two groups of cells. The default is 0.25. Increasing `logfc.threshold` speeds up the function but can miss weaker signals.

-   `min.pct` Only test genes that are detected in a minimum fraction of cells in either of the two populations. It is meant to speed up the function by not testing infrequently expressed genes. The default is 0.1.

-   `only.pos` Only return positive markers (FALSE by default).

One may change the arguments depending on how stringent to be.

We look at the marker genes in the clusters in sample "treated".

```{r}
treated.markers <- FindAllMarkers(subset(filtered, subset = Sample_Name == "SampleTag02_hs"), 
                                  assay = "RNA",
                                  only.pos = TRUE, 
                                  min.pct = 0.25, 
                                  logfc.threshold = 0.25)
```

Let's look at the top 2 significant marker genes in each cluster.

```{r}
treated.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)
```

We get a table with the information on marker genes.

`p_val` P-value before multiple testing.

`avg_log2FC` Log2 fold change on the average gene expression between one cluster and the rest of cluster combined.

`pct.1` Percentage of cells expressing the gene in one group. Here is one cluster.

`pct.2` Percentage of cells expressing the gene in another group. Here is the rest of cluster combined.

`p_val_adj` p-value with multiple testing correction.

`gene` Gene name.

## 1.2 `FindMarkers`

We can use the `FindMarkers` to determine the differentially expressed genes between two clusters.

```{r}
example_findmarkers <- FindMarkers(filtered,
                          assay = "RNA",
                          ident.1 = 1,
                          ident.2 = 2) %>% 
  rownames_to_column("gene")
```

# 2. Automatic annotation

So far, we already look at the marker genes based on knowledge (manual annotation) and between clusters by differential expression analysis. The manual annotation usually takes time, and we might need to go back and forth to annotate the clusters. Another way to annotate the clusters is through the automatic annotation.

## 2.1 Reference-based

We look at the publicly available dataset with annotation and build a reference set based on its information. Then we can "transfer" the annotation from the reference to the query. Seurat provides the interface to connect with `Azimuth`.

::: callout-note
Because the annotation information is based on the reference set, the reference ideally is similar to the query. They may come from the same tissue, same organism, similar experimental procedure, etc.
:::

```{r}
filtered <- Azimuth::RunAzimuth(filtered, reference = "pbmcref", verbose = F)
```

We can look at the automatic annotation on UMAP.

```{r}
p1 <- DimPlot(filtered, reduction = "umap", 
              group.by = "predicted.celltype.l1", 
              label = TRUE, label.size = 3 ,repel = TRUE) + 
  NoLegend()

p2 <- DimPlot(filtered, reduction = "umap", 
              group.by = "RNA_snn_res.0.4", 
              label = TRUE, label.size = 3 ,repel = TRUE) + 
  NoLegend()

p1|p2
```

## 2.1 Marker-based

We use ScType (<https://github.com/IanevskiAleksandr/sc-type/blob/master/README.md>) to annotate cell population based on known immune markers.

ScType also has an open source portal: <https://sctype.app/>.

```{r}
# # load ScType R functions
# source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R"); source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
# 
# # get cell-type-specific gene sets from our in-built database (DB)
# # list downloaded from https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_short.xlsx
# # e.g. Immune system, Liver, Pancreas, Kidney, Eye, Brain
# gs_list <-  gene_sets_prepare("data/ScTypeDB_short.xlsx", "Immune system") 
# 
# # assign cell types
# # NOTE: scRNAseqData parameter should correspond to your input scRNA-seq matrix. 
# # In case Seurat is used, it is either GetAssayData(seuratObj[["RNA"]], layer = "scale.data"), in case sctransform is used for normalization,
# # or GetAssayData(seuratObj[["integrated"]], layer = "scale.data"), in case a joint analysis of multiple single-cell datasets is performed.
# es.max <-  sctype_score(scRNAseqData = GetAssayData(filtered[["RNA"]], layer = "scale.data"), 
#                         scaled = TRUE, 
#                         gs = gs_list$gs_positive, 
#                         gs2 = gs_list$gs_negative)
# 
# # merge by cluster
# cL_resutls = do.call("rbind", lapply(unique(filtered@meta.data$RNA_snn_res.0.4), function(cl){
#     es.max.cl = sort(rowSums(es.max[ ,rownames(filtered@meta.data[filtered@meta.data$RNA_snn_res.0.4==cl, ])]), decreasing = !0)
#     head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(filtered@meta.data$filtered==cl)), 10)
# }))
# 
# sctype_scores = cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  
# 
# # set low-confident (low ScType score) clusters to "unknown"
# sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] = "Unknown"
# ```
# 
# ```{r}
# filtered@meta.data$customclassif = ""
# 
# for(j in unique(sctype_scores$cluster)){
#   cl_type = sctype_scores[sctype_scores$cluster==j,]; 
#   filtered@meta.data$customclassif[filtered@meta.data$RNA_snn_res.0.4 == j] = as.character(cl_type$type[1])
# }
# 
# DimPlot(filtered, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'customclassif')   
```

```{r}
#| echo: false

# save the seurat object
save.image(file = "data/chapter5.RData")
```
