---
title: "Chapter 6 Differential Expression Analysis"
execute:
  warning: false
  message: false
format:
  html:
    embed-resources: true
---

```{r}
#| echo: false

# Load RData
load("data/chapter5.RData")
```

```{r}
# libraries for this chapter
library(tidyverse)
library(Seurat)
library(data.table)
library(ggpubr)
```

# 1. FindMarkers

## 1.1 DE analysis across conditions

In the last chapter, we used `FindMarkers` to find the marker genes between two clusters by specifying `ident.1` and `ident.2`.

Changing the `ident.1` and `ident.2` arguments, we can perform differential expression analysis between two groups.

For instance, we want to find genes deferentially expressed between naïve and treated PBMC.

```{r}
# change default assay to "RNA"
DefaultAssay(filtered) <- "RNA"

# change default identity to "Sample_Name"
Idents(filtered) <- "Sample_Name"

# Find genes differentially expressed
# SampleTag02_hs: treated, SampleTag01_hs: naive
treated_naive <- 
  FindMarkers(filtered, ident.1 = "SampleTag02_hs", ident.2 = "SampleTag01_hs") %>% 
  rownames_to_column("gene") %>% 
  dplyr::filter(p_val_adj < 0.05)
```

We can compare the gene expression between two groups by specifying `ident.1` and `ident.2`.

```{r}
head(treated_naive)
```

-   `gene` Gene symbol.

-   `p_val` P-value without multiple testing.

-   `avg_log2FC` log2 fold-change of the average expression between the two groups.

-   `pct.1` Percentage of cells in ident.1 expressing the gene.

-   `pct.2` Percentage of cells in ident.2 expressing the gene.

-   `p_val_adj` Adjusted p-value.

## 1.2 Perform DE analysis within the same cell type across conditions

We can also perform DE analysis within the same cell type across conditions.

```{r}
filtered$predicted.celltype.l1.tag <- paste(filtered$predicted.celltype.l1, filtered$Sample_Name, sep = "_")

# Find genes differentially expressed between activated vs. naive samples in CD4 T cells
Idents(filtered) <- "predicted.celltype.l1.tag"

cd4t.deg <- FindMarkers(filtered, 
                       ident.1 = "CD4 T_SampleTag02_hs", 
                       ident.2 = "CD4 T_SampleTag01_hs", 
                       verbose = FALSE)

head(cd4t.deg, n = 10)
```

# 2. Pseudobulk analysis

## 2.1 Setup

Let's assume we have three biological replicates in each condition (we make them up for the purpose of illustration).

```{r}
# make a fake group information
filtered$donor <- sample(x = c('A', 'B', 'C'), size = ncol(filtered), replace = TRUE)

pseudo <- AggregateExpression(filtered, assays = "RNA", return.seurat = T, 
                              group.by = c("Sample_Name", "donor", "predicted.celltype.l1"))

# perform pseudobulk
pseudo$predicted.celltype.l1.tag <- paste(pseudo$predicted.celltype.l1, pseudo$Sample_Name, sep = "_")

Idents(pseudo) <- "predicted.celltype.l1.tag"

cd4t.bulk <- FindMarkers(object = pseudo, 
                         ident.1 = "CD4 T_SampleTag02-hs", 
                         ident.2 = "CD4 T_SampleTag01-hs",
                         test.use = "DESeq2")

head(cd4t.bulk, n = 15)
```

```{r}
#| echo: false

# save the seurat object
save.image(file = "data/chapter6.RData")
```
