---
title: "Chapter 7 Differential Expression Analysis"
execute:
  warning: false
  message: false
---

```{r}
# libraries for this chapter
library(tidyverse)
library(Seurat)
library(data.table)
library(ggpubr)
```

# Load RData

Load the RData which contains all files from the previous chapter.

```{r}
load("data/intermediate_data/chapter6.RData")
```

# 1. FindMarkers

## 1.1 DE analysis across conditions

In the last chapter, we used `FindMarkers` to find the marker genes between two clusters by specifying `ident.1` and `ident.2`.

Changing the `ident.1` and `ident.2` arguments, we can perform differential expression analysis between two groups.

For instance, we want to find genes deferentially expressed between naïve and treated PBMC.

```{r}
# assign the sample identity that we want to work on different sample groups
Idents(integrate.filtered.rna) <- "Sample_Name"

# change default assay to "RNA"
DefaultAssay(integrate.filtered.rna) <- "RNA"

# Find genes differentially expressed
treated_naive <- 
  FindMarkers(integrate.filtered.rna, ident.1 = "treated", ident.2 = "naive") %>% 
  rownames_to_column("gene") %>% 
  dplyr::filter(p_val_adj < 0.05)
```

We can compare the gene expression between two groups by specifying `ident.1` and `ident.2`.

```{r}
head(treated_naive)
```

-   `gene` Gene symbol.

-   `p_val` P-value without multiple testing.

-   `avg_log2FC` log2 fold-change of the average expression between the two groups.

## 1.2 Perform DE analysis within the same cell type across conditions

We can also perform DE analysis within the same cell type across conditions.

```{r}
# subset CD4 T cells
integrate.filtered.rna$celltype.stim <- paste(integrate.filtered.rna$Sample_Name, integrate.filtered.rna$predicted.celltype.l1, sep = "_")

# Find genes differentially expressed between activated vs. naive samples in CD4 T cells
Idents(integrate.filtered.rna) <- "celltype.stim"

cd4t.sc <- FindMarkers(integrate.filtered.rna, 
                       ident.1 = "treated_CD4 T", 
                       ident.2 = "naive_CD4 T", 
                       verbose = FALSE)

head(cd4t.sc, n = 10)
```

# 2. Pseudobulk analysis

## 2.1 Setup

Let's assume we have three biological replicates in each condition (we make them up for the purpose of illustration).

```{r}
# make a fake group information
integrate.filtered.rna$group <- sample(x = c('A', 'B', 'C'), size = ncol(integrate.filtered.rna), replace = TRUE)

# pseudo-bulk by per condition per cell type
cd4t.bulk <- Seurat:::PseudobulkExpression(integrate.filtered.rna, 
                                        assays = "RNA",
                                     pb.method = 'aggregate',
                                     group.by = c('Sample_Name', 'predicted.celltype.l1', 'group'),
                                     slot = "counts",
                                     return.seurat = T)
  
# tidy up names
cd4t.bulk$Sample_Name <- cd4t.bulk %>% colnames() %>% 
  str_split(., pattern = "_") %>% 
  lapply(., "[", 1) %>% 
  unlist()

cd4t.bulk$cellType <- cd4t.bulk %>% colnames() %>% 
  str_split(., pattern = "_") %>% 
  lapply(., "[", 2) %>% 
  unlist

cd4t.bulk$celltype.stim <- paste(cd4t.bulk$Sample_Name, cd4t.bulk$cellType, sep = "_")

# perform pseudobulk
Idents(cd4t.bulk) <- "celltype.stim"

cd4t.bulk <- FindMarkers(cd4t.bulk, 
                         ident.1 = "treated_CD4 T", 
                         ident.2 = "naive_CD4 T", 
                         test.use = "DESeq2")
```

```{r}
# compare the DE P-values between the single-cell level and the pseudobulk level results
cd4t.bulk <- cd4t.bulk %>% 
  `colnames<-`(paste0(names(cd4t.bulk), ".bulk")) %>% 
  mutate(gene = rownames(cd4t.bulk))

cd4t.sc <- cd4t.sc %>% 
  `colnames<-`(paste0(names(cd4t.sc), ".sc")) %>% 
  mutate(gene = rownames(cd4t.sc))

merge_dat <- merge(cd4t.bulk, cd4t.sc, by = "gene") %>% 
  arrange(p_val.bulk)

# Number of genes that are marginally significant in both; marginally significant only in bulk; and marginally significant only in single-cell
common <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 & 
                                merge_dat$p_val.sc < 0.05)]
only_sc <- merge_dat$gene[which(merge_dat$p_val.bulk > 0.05 & 
                                  merge_dat$p_val.sc < 0.05)]
only_bulk <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 & 
                                    merge_dat$p_val.sc > 0.05)]
```

The majority of DEGs are shared in both pseudobulk and single-cell analysis. The log2FC is also correlated.

```{r}
ggscatter(merge_dat, x = "avg_log2FC.bulk", y = "avg_log2FC.sc", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson")
```

# 3. Save

Save the files into RData for documentation.

```{r}
# save.image(file = "data/intermediate_data/chapter7.RData")
```
