---
title: "Chapter 8 Enrichment Analysis"
execute:
  warning: false
  message: false
---

```{r}
# libraries for this chapter
library(tidyverse)
library(Seurat)
library(data.table)
library(org.Hs.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(enrichplot)
library(biomaRt)
```

# Load RData

Load the RData which contains all files from the previous chapter.

```{r}
load("data/intermediate_data/chapter7.RData")
```

Oftentimes, enrichment analysis takes the differentially expressed genes and looks at which gene ontology (GO) terms are enriched.

GO terms defines concepts/classes used to describe gene function and relationships between these concepts. `clusterProfiler` is an R package implementing many methods to analyze and visualize functional profiles from a group of features. It classifies functions in three aspects.

-   MF: **M**olecular **F**unction: molecular activities of gene products.

-   CC: **C**ellular **C**omponent: where gene products are active.

-   BP: **B**iological **P**rocess: pathways and larger processes made up of the activities of multiple gene products.

This chapter uses the differentially expressed genes in CD4+ T cells from last chapter as an example.

# 1. Get Entrez ID

`clusterProfiler` prefers the input of Entrez ID. We use `biomaRt` package to convert the gene symbol into Entrez ID.

::: callout-note
`biomaRt` is handy to get information about features. You can use it to get gene symbols, genomic coordinates, transcript sequences, etc. You can explore what data is stored in `biomaRt` with `listAttributes`.
:::

The differentially expressed genes in CD4+ T cells between naïve and treated PBMCs are stored in the matrix `cd4t`. The gene symbol in this list can be converted into Entrez ID by `biomaRt`.

Firstly, connect to the Ensembl database and specify we want to retrieve information from the dataset of Homo Sapien.

```{r}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
```

-   `biomart` BioMart database connect to. Here, we want to connect to `ensembl` database.

-   `dataset` Dataset to use. Here specify `hsapiens_gene_ensembl`.

Next, we can give a list of gene symbols and retrieve the Entrez ID through `getBM`.

```{r}
results <- getBM(attributes = c("entrezgene_id", "external_gene_name"),
                 filters = "external_gene_name", 
                 values = cd4t$gene,
                 mart = mart)
```

-   `attributes` Attributes to retrieve. We want to have the gene symbols and Entrez ID in the results.

-   `filters` Which filter to use to filter the dataset.

-   `values` A list of features as a query.

-   `mart` Object of class Mart, created with the `useMart` function.

A table of two columns is created and stored in the matrix of `results`.

```{r}
results %>% head
```

Not every gene has the Entrez ID. We keep only genes that have a matched Entrez ID.

```{r}
cd4t <- left_join(cd4t, results, by = c("gene" = "external_gene_name")) %>% 
  filter(!is.na(entrezgene_id))
```

# 2. GO over-representation analysis

Over Representation Analysis (ORA) is a widely used approach to determine whether known biological functions or processes are over-represented (= enriched) in an experimentally-derived gene list, e.g. a list of differentially expressed genes (DEGs).

We use the `enrichGO` to get the enriched GO terms in the list of DEGs.

```{r}
cd4t.sig <- dplyr::filter(cd4t, p_adj.loc < 0.05, logFC > 1) %>% 
  arrange(p_adj.loc)

# get the enriched GO terms
ego <- enrichGO(gene = cd4t.sig$entrezgene_id,
                keyType = "ENTREZID",
                OrgDb = org.Hs.eg.db,
                ont = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.01,
                qvalueCutoff = 0.05,
                readable = TRUE)
```

-   `gene` A list of Entrez gene id.

-   `keyType` The keytype of input gene.

-   `OrgDb` The database with detailed information about the species.

-   `org.Hs.eg.db` Specify the database of Homo Sapien.

-   `ont` One of "BP" (biological process), "MF" (molecular function), and "CC" (cellular component) subontologies, or "ALL" for all three.

-   `pAdjustMethod` Multiple testing method for p-value.

-   `pvalueCutoff` Adjusted pvalue cutoff on enrichment tests to report.

-   `qvalueCutoff` Tge qvalue cutoff on enrichment tests to report as significant.

-   `readable` Whether mapping gene ID to gene Name.

Let's convert the `ego` object into a data frame and visualize the results in a bar plot.

```{r}
ego.table <- as.data.frame(ego)

head(ego.table)
```

-   `GeneRatio` The gene counts from the input list found in the gene set of a GO term / gene counts of input list.

-   `BgRatio` The gene counts of a geneset / size of all of the unique genes in the collection of genesets.

```{r}
barplot(ego, showCategory = 18)
```

# 3. Gene set enrichment analysis

All genes can be used as input of gene set enrichment analysis (GSEA). We first rank the genes based on `stat`.

```{r}
# order genes based on stat
cd4t <- cd4t[order(-cd4t$stat), ]

gene_list <- cd4t$stat
names(gene_list) <- cd4t$entrezgene_id
```

We use `gseGO` to perform GSEA and convert the results into a dataframe.

```{r}
# perform GSEA
gse <- gseGO(gene_list,
             keyType = "ENTREZID",
             ont = "BP",
             OrgDb = "org.Hs.eg.db")

# convert the results into a dataframe
gse.table <- as.data.frame(gse) %>% 
  mutate(geneSetID = 1:dim(.)[1])
```

`enrichmentScore` The enrichment score (ES) represents the degree to which a set S is over-represented at the top or bottom of the ranked list L.

We can look at the enrichment of some GO terms in GSEA plot.

```{r}
gseaplot2(gse, geneSetID = c(188, 194))
```

We have significant enriched term in the unregulated genes. Along the x axis is the ranked genes. Each of the vertical line is one of those genes that are found in this gene set. Along the running line, you can see a lot of genes running together and it falls out when the gene becomes more sparse in the list.

# 4. Save

Save the files into RData for documentation.

```{r}
save(integrate.filtered.rna, expMat, cd4t, ego.table, gse.table, 
     file = "data/intermediate_data/chapter8.RData")
```
